\textls[-0]{\thetitle}

### 2023-03-21 metatheory-condensed.tex
comment
add more details about bold-P

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT %(predicative) MLTT
(and \cref{fig:fmltt-selected} \TT).
Dependent function types \nTyPi{x}{A}{B},
dependent pair types \nTySigma{x}{A}{B}, and
identity types \TyId{t_1}{t_2} are standard.
So are their introduction and elimination forms.
Capture-avoiding substitution is notated $\bullet[\bullet/x]$.
A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{tm/s})~\cite{aspinall1995singleton,stone2000}.
Definitional equalities have the forms %(for each judgement)
\goodCtx{\Gm_1 \equiv \Gm_2}{},
\goodType{\Gm}{T_1 \equiv T_2}{},
\goodTerm{\Gm}{t_1 \equiv t_2}{T},
etc.
Following \citet{altkap2016}, we regard our (intrinsically typed) syntax
as being quotiented by these equalities.
Quotienting facilitates coercion along equalities---given
\goodCtx{\Gm_1 \equiv \Gm_2}{} and
\goodType{\Gm_1}{T_1 \equiv T_2}{},
the derivation of \goodTerm{\Gm_1}{t}{T_1} is considered definitionally equal to
a derivation of \goodTerm{\Gm_2}{t}{T_2}. %\YZ{Is this accurate??}\EDJreply{Yes. In QIIT model, it looks like `t : Tm G1 T1' and  `t : Tm G2 T2'. In extenional Agda, because ``G1 = G2'', ``T1 = T2'', thus the two types `Tm G2 T2 = Tm G1 T1'. So they are the same derivation in QIIT. In our setup, we have the same idea, because after quotient, G1 refers to same object as G2, so the derivation is the same.}
%\YZ{should there be a convertibility rule?}\EDJreply{in this style, the quotient by judgemental equality is applied, so A is equal to B even in meta-logic, which means t : A is the same as t : B. Maybe this is something require emphasis? Because it is usually handled by LF (like QIIT) but now we are using bare hands so.... maybe we need to emphasize this}
%For each derivation (of each judgment)\YZ{I don't get what 'each derivation' or 'each judgment' means}\EDJreply{(G |- t : T), which looks like (t : Tm G T) in QIIT/Agda is the derivation; so basically the term is the derivation (or proof), the types (Tm G T) is the judgement. It applies to each kind of judgement, for example, (G |- ), which looks like (G : Con), (G |-) is the derivation and (_ |- )is the judgement \\ 
%Note that without referring to QIIT/Agda, the definition of judgement and derivation is still clear -- (G |- _ : T) is a judgement and a concrete proof/derivation we get for t like (G |- t : T) is the derivation. Similar for ( _ |- ) vs. (G |-) and (G |- _ ) vs (G |- T). \\ 
%I think in intrinsic typing, knowing t and knowing (G |- t : T) is exactly the same (i.e. there are no two different derivations lead to the same t; for each t, there is only one derivation for it). This is not true for extrinsic typing. \\
%If this prequisite knowledge is too involved, feel free to delete this sentence, I think I just want to say there are equality/quotient everywhere between all kinds of derivations(proof?data?) -- equality between terms, types, substitutions, contexts, linkage signatures ...\\
%I think this is something standard from Martin Lof, since Martin Lof is the first guy talk about judgement.
%}

@@

Well-formed \emph{linkage signatures} have the judgment form
$\goodSig{\Gm}{\lsig}{n}$.
As rules \ruleref{lsig/empty} and \ruleref{lsig/add} show, a linkage
signature is a list of $n$~types.
%
Well-formed \emph{linkages} have the judgment form $\goodTerm{\Gm}{\lkg}{\TyLkg{\lsig}}$,
where $\TyLkg{\lsig}$ is a type formed by~$\lsig$.
As rules \ruleref{l/empty} and \ruleref{l/add} show, a linkage is a list
of $n$~terms, each representing a field of the family modeled by the linkage.

In rule \ruleref{l/add}, the second premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the context of the current field~$t$, controlling
how the types of the fields prior to~$t$ are exposed to the typing of~$t$.
Crucially, %unlike the second field in a dependent pair (\ruleref{tm/pair}),
the premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ makes clear that
the typing of~$t$ in \ruleref{l/add} is \emph{universally} quantified---%
rather than \emph{existentially} quantified as is in \ruleref{tm/pair}---%
over how the fields in $t$'s context are defined.
Late binding enables reuse.
A different linkage~$\ell'$ that overrides fields in $t$'s context (and thus models a derived family)
can reuse~$t$---without retyping it---by first
projecting~$t$ from $\LkgAdd{\lkg}{t}$ and then appending it to~$\lkg'$.
%\EDJ{I suggest change this sentence to `by projecting it out of parent linkage and adding it in the derived linkage'}.

@@

We use \cref{fig:stlc-linkage-typing} to illustrate.
On the left is code excerpted from \cref{fig:stlc-mechanized}, and on the right
is how the corresponding fields are modeled and typed in \TT.
%
$\lkg_i$ is the linkage that adds the $i$-th field~$t_i$ with the typing
$\goodTerm{%
\tikzmarkin[disable rounded corners=true,set fill color=A-color,set border color=A-color]{A-i}(-0.02,-0.10)(-0.02,0.28)
\mathit{self}_i : A_i%
\tikzmarkend{A-i}%
}{%
\tikzmarkin[disable rounded corners=true,set fill color=t-color,set border color=t-color]{t-i}(-0.02,-0.10)(-0.02,0.28)%
t_i%
\tikzmarkend{t-i}%
}{%
\tikzmarkin[disable rounded corners=true,set fill color=T-color,set border color=T-color]{T-i}(-0.02,-0.10)(-0.02,0.28)%
T_i%
\tikzmarkend{T-i}%
}$.
%
The context types $A_i$ are a dependent tuple type, but for readability,
we write them as dependent record types that give labels to the fields.
%we use a notation with explicit names
%in this typing judgment:
%we name the context as $\mathit{self}$ (cf.~\cref{fig:stlc-compiled})
%and write its type~$A_i$, which is a dependent tuple type, as a
%dependent record type with labels.
%
\begin{itemize}
% [leftmargin=4ex,itemsep=.5ex]
  [labelsep=*,leftmargin=1pc,itemsep=3pt]

\item 
The first field $t_1$ is defined as $\wcode{\wsig_{\texttt{tm}}}$
and given the singleton type $\TyS{\wcode{\wsig_{\texttt{tm}}}}$, where $\wsig_{\texttt{tm}}$
is the W-type signature for \lsti{tm}.
This typing is recorded by $\lsig_1$ and is thus available in all $\lsig_i$'s.

@@

\item
Similarly, $\wsig_{\texttt{tm}}$ is exposed in the typing context of
$t_{10}$, which models the recursive function \lsti{subst} by invoking
the recursor $\Wrec{\wsig_{\texttt{tm}}}{\cDot}{\cDot}$.
%
Like \lsti{subst}, partial recursors in \Lang are axiomatized by the plugin
(\cref{sec:latebind,sec:prec}), and they can similarly be defined in \TT
using $\texttt{Wrec}$.

@@

The third premise
$\goodTerm{\Gm,x : \TyTkg{\lsig}}{\seal}{A}$
in rules \ruleref{l/add} and \ruleref{lsig/add} is
responsible for hiding W-type signatures.
Here, $\TyTkg{\lsig}$ is a dependent tuple type (rule \ruleref{tyeq/pk/add})
that packages the types of all fields preceding the current field~$t$,
and $\seal$ is the term turning the tuple $\Tkg{\lkg}$ of type
$\TyTkg{\lsig}$ (rule \ruleref{tmeq/pk/add}) into a new tuple of type~$A$
that hides W-type signatures behind $\cU$, if necessary.
%
It is straightforward to find the~$\seal$ that fits the bill, though this process
is not automated in \TT.
In particular, when no hiding is needed, $\seal$ is $\var{0}$
%\EDJ{$\var{0}$ is not identity (substitution), but the first variable, right? Or you meant to say it is identity ``sealing''?}\YZreply{Yes, I meant identity sealing. I cannot say 'sealing' because the paper currently does not use the word sealing anywhere.}\EDJreply{What about just remove the parenthesis and ``the identity''}
and $A$ is $\TyTkg{\lsig}$.
In \cref{fig:stlc-linkage-typing}, $\seal_1$, $\seal_2$, and $\seal_{10}$ are $\var{0}$,
and $\seal_6$ is from
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \TyS{\wcode{\wsig_{\texttt{tm}}}},\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$
\endgroup
to
\begingroup
\fontsize{9}{9}\selectfont%
$\big[\texttt{tm} : \cU,\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]$%
\endgroup
.

When a family is concluded (e.g., \lsti!End!~\lsti!STLC!),
a linkage~$\lkg$ containing all the fields is available
(e.g., $\lkg_{53}$ in \cref{fig:stlc-linkage-typing}).
Fields of the family can then be accessed by projecting them out of the tuple
$\Tkg{\lkg}$.

>>>>

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT %(predicative) MLTT
(and \cref{fig:fmltt-selected} \TT).
Dependent function types \nTyPi{x}{A}{B},
dependent pair types \nTySigma{x}{A}{B}, and
identity types \TyId{t_1}{t_2} are standard.
So are their introduction and elimination forms.
Capture-avoiding substitution is notated $\bullet[\bullet/x]$.
We use $x$ and $\mathit{self}$ to denote variables.
A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{tm/s})~\cite{aspinall1995singleton,stone2000}.
Definitional equalities have the forms %(for each judgement)
\goodCtx{\Gm_1 \equiv \Gm_2}{},
\goodType{\Gm}{T_1 \equiv T_2}{},
\goodTerm{\Gm}{t_1 \equiv t_2}{T},
etc.
Following \citet{altkap2016}, we regard our (intrinsically typed) syntax
as being quotiented by these equalities.
Quotienting facilitates coercion along equalities---given
\goodCtx{\Gm_1 \equiv \Gm_2}{} and
\goodType{\Gm_1}{T_1 \equiv T_2}{},
the derivation of \goodTerm{\Gm_1}{t}{T_1} is considered definitionally equal to
a derivation of \goodTerm{\Gm_2}{t}{T_2}. %\YZ{Is this accurate??}\EDJreply{Yes. In QIIT model, it looks like `t : Tm G1 T1' and  `t : Tm G2 T2'. In extenional Agda, because ``G1 = G2'', ``T1 = T2'', thus the two types `Tm G2 T2 = Tm G1 T1'. So they are the same derivation in QIIT. In our setup, we have the same idea, because after quotient, G1 refers to same object as G2, so the derivation is the same.}
%\YZ{should there be a convertibility rule?}\EDJreply{in this style, the quotient by judgemental equality is applied, so A is equal to B even in meta-logic, which means t : A is the same as t : B. Maybe this is something require emphasis? Because it is usually handled by LF (like QIIT) but now we are using bare hands so.... maybe we need to emphasize this}
%For each derivation (of each judgment)\YZ{I don't get what 'each derivation' or 'each judgment' means}\EDJreply{(G |- t : T), which looks like (t : Tm G T) in QIIT/Agda is the derivation; so basically the term is the derivation (or proof), the types (Tm G T) is the judgement. It applies to each kind of judgement, for example, (G |- ), which looks like (G : Con), (G |-) is the derivation and (_ |- )is the judgement \\ 
%Note that without referring to QIIT/Agda, the definition of judgement and derivation is still clear -- (G |- _ : T) is a judgement and a concrete proof/derivation we get for t like (G |- t : T) is the derivation. Similar for ( _ |- ) vs. (G |-) and (G |- _ ) vs (G |- T). \\ 
%I think in intrinsic typing, knowing t and knowing (G |- t : T) is exactly the same (i.e. there are no two different derivations lead to the same t; for each t, there is only one derivation for it). This is not true for extrinsic typing. \\
%If this prequisite knowledge is too involved, feel free to delete this sentence, I think I just want to say there are equality/quotient everywhere between all kinds of derivations(proof?data?) -- equality between terms, types, substitutions, contexts, linkage signatures ...\\
%I think this is something standard from Martin Lof, since Martin Lof is the first guy talk about judgement.
%}

@@

\emph{Linkage signatures} have the form $\goodSig{\Gm}{\lsig}{n}$.
As rules \ruleref{lsig/empty} and \ruleref{lsig/add} show, a linkage
signature is a list of $n$~types.
%
\emph{Linkages} have the form $\goodTerm{\Gm}{\lkg}{\TyLkg{\lsig}}$,
where $\TyLkg{\lsig}$ is a type formed by~$\lsig$.
As rules \ruleref{l/empty} and \ruleref{l/add} show, a linkage is a list
of $n$~terms, each representing a field of the family modeled by the linkage.

In rules \ruleref{l/add} and \ruleref{lsig/add},
the second premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the context of the current field~$t$, controlling
how the types of the fields prior to~$t$ are exposed to the typing of~$t$.
Crucially, %unlike the second field in a dependent pair (\ruleref{tm/pair}),
the premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ makes clear that
the typing of~$t$ in \ruleref{l/add} is \emph{universally} quantified---%
rather than \emph{existentially} quantified as is in \ruleref{tm/pair}---%
over how the fields in $t$'s context are defined.
Late binding enables reuse.
A different linkage~$\ell'$ that overrides fields in $t$'s context (and thus models a derived family)
can reuse~$t$---without retyping it---by first
projecting~$t$ from $\LkgAdd{\lkg}{t}$ and then appending it to~$\lkg'$.
%\EDJ{I suggest change this sentence to `by projecting it out of parent linkage and adding it in the derived linkage'}.

@@

We use \cref{fig:stlc-linkage-typing} to illustrate.
On the left is code excerpted from \cref{fig:stlc-mechanized}, and on the right
is how the corresponding fields are modeled and typed in \TT.
%
$\lkg_i$ is the linkage that adds the $i$-th field~$t_i$ with the typing
$\goodTerm{%
\tikzmarkin[disable rounded corners=true,set fill color=A-color,set border color=A-color]{A-i}(-0.02,-0.10)(-0.02,0.28)
\mathit{self} : A_i%
\tikzmarkend{A-i}%
}{%
\tikzmarkin[disable rounded corners=true,set fill color=t-color,set border color=t-color]{t-i}(-0.02,-0.10)(-0.02,0.28)%
t_i%
\tikzmarkend{t-i}%
}{%
\tikzmarkin[disable rounded corners=true,set fill color=T-color,set border color=T-color]{T-i}(-0.02,-0.10)(-0.02,0.28)%
T_i%
\tikzmarkend{T-i}%
}$.
%
The context types $A_i$ are a dependent tuple type, but for readability,
we write them as dependent record types that give labels to the fields.
%we use a notation with explicit names
%in this typing judgment:
%we name the context as $\mathit{self}$ (cf.~\cref{fig:stlc-compiled})
%and write its type~$A_i$, which is a dependent tuple type, as a
%dependent record type with labels.
%
\begin{itemize}
% [leftmargin=4ex,itemsep=.5ex]
  [labelsep=*,leftmargin=1pc,itemsep=3pt]

\item 
The first field $t_1$ is defined as $\wcode{\wsig_{\texttt{tm}}}$
and given the singleton type $\TyS{\wcode{\wsig_{\texttt{tm}}}}$, where $\wsig_{\texttt{tm}}$
is the W-type signature constructed earlier for the extensible inductive type~\lsti{tm}.
This typing is recorded by $\lsig_1$ and is thus available in all $\lsig_i$'s.

@@

\item
Likewise, $\wsig_{\texttt{tm}}$ is exposed in the typing context of
$t_{10}$, which models the recursive function \lsti{subst} by invoking
the recursor $\Wrec{\wsig_{\texttt{tm}}}{\cDot}{\cDot}$.
%
Like \lsti{subst}, partial recursors in \Lang are axiomatized by the plugin
(\cref{sec:latebind,sec:prec}), and they can similarly be defined in \TT
using $\texttt{Wrec}$.

@@

In rules \ruleref{l/add} and \ruleref{lsig/add},
the third premise $\goodTerm{\Gm,x : \TyTkg{\lsig}}{\seal}{A}$ is
responsible for hiding W-type signatures.
Here, $\TyTkg{\lsig}$ packages $\lsig$---recall that $\lsig$ contains the
(self-parameterized) types of all the fields preceding the current
field~$t$---into a dependent tuple type (\ruleref{tyeq/pk/add}).
The term $\seal$ turns a tuple of type $\TyTkg{\lsig}$ into a new tuple of type~$A$
that hides W-type signatures behind $\cU$, if necessary.
%
It is straightforward to find the~$\seal$ that fits the bill, though this process
is not automated in \TT.
In particular, when no hiding is needed, $\seal$ is simply~$x$,
%\EDJ{$\var{0}$ is not identity (substitution), but the first variable, right? Or you meant to say it is identity ``sealing''?}\YZreply{Yes, I meant identity sealing. I cannot say 'sealing' because the paper currently does not use the word sealing anywhere.}\EDJreply{What about just remove the parenthesis and ``the identity''}
and $A$ is $\TyTkg{\lsig}$.
In \cref{fig:stlc-linkage-typing}, $\seal_1$, $\seal_2$, and $\seal_{10}$ are simply $x$,
and $\seal_6$ has the typing
\begingroup
\fontsize{9}{9}\selectfont%
\goodTerm{
  \mathit{self} :
  \big[\texttt{tm} : \TyS{\wcode{\wsig_{\texttt{tm}}}},\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]
}{
  \seal_6
}{
  \big[\texttt{tm} : \cU,\ \texttt{tm\_unit} : \El{\texttt{tm}},\ \dadada\big]
}.
\endgroup

When a family is concluded (e.g., \lsti!End!~\lsti!STLC!),
a linkage~$\lkg$ containing all the fields is available
(e.g., $\lkg_{53}$ in \cref{fig:stlc-linkage-typing}).
Fields of the family can then be accessed by projecting them out of the tuple
$\Tkg{\lkg}$.
As \ruleref{tmeq/pk/add} indicates, $\Tkg{\lkg}$ ties the recursive knot:
it packages the linkage~$\lkg$ into a dependent tuple of type~$\TyTkg{\lsig}$,
by instantiating the $\mathit{self}$ parameters.
